---
title: "CRN compliant GHCN stations"
author: "Steven Mosher"
date: "January 6, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Classifying GHCN via CRN  metadata metrics

 Given that we have CRN station characteristics we can cull the GHCN V4 stations using the values
 from CRN. That is select GHCN stations that have similar metadata for the local area. Precise site
 level metrics can also be assessed given higher resultion metadata and precise site location. Here we leave 
 problem for a future excerise. The goal here is to select CRN like candidates.

```{r, echo=FALSE,message=FALSE}
  source("libraries.R")
  source("filenames.R")
  GHCN <- tbl_df(read.csv(ghcnmeta, stringsAsFactors = F))
  CRN <- tbl_df(read.csv(crnmeta, stringsAsFactors = F))
```

 

## Local Area Covered by Water

For every site we calculate the percentage of water within a 10km radius using the ESA 300meter Landcover product

```{r, message = FALSE,warning=FALSE}
f2 <- ggplot(GHCN, aes(100* WaterArea/314))+geom_histogram() +
      ggtitle("Percentage Water Cover",subtitle = "10 km radius")+
      xlab("Percentage") + ylab("Station Count")
f2
```

## Local Urban Area

Using CRN as a template we select those GHCN stations that have a comparable amount of Urban area within
the 10km boundary.The limit is set at the 90th percentile of CRN values

```{r, message = FALSE,warning=FALSE}
p <- quantile(CRN$UrbanArea10K,probs=seq(0,1,by=.05))[19]
Maxurban <- p
GHCN     <- GHCN %>% mutate(Umax = ifelse(UrbanArea10K < Maxurban,TRUE,FALSE))
f1       <- ggplot(GHCN,aes(Umax, fill=Umax))+geom_bar()+
            ggtitle("GHCN Stations matching CRN",subtitle = "10 km urban area")+
            xlab("CRN-like") + ylab("Station Count")
f1
f2       <- ggplot(filter(GHCN,Umax==TRUE), aes(100* UrbanArea10K/314))+geom_histogram() +
           ggtitle("Percentage Urban Cover",subtitle = "10 km radius")+
           xlab("Percentage") + ylab("Station Count")
      
f2
```

The next criteria we apply os the local area Population, using the GPW density figure from CRN to
limit the stations. In addition we retain the sites were there is no population data as these are
typically ship and small island locations.
 

To get a visual sense of the amount of urbanity these figures represent we look at a google map visualization and select stations with the highest amount of urban area 

```{r, message = FALSE,warning=FALSE}
  Maxpop <- quantile(CRN$GPW10km_15_Density,probs=seq(0,1, by=.05))[19]
GHCN     <- GHCN %>% mutate(Pmax = ifelse(GPW10km_15_Density < Maxpop | is.na(GPW10km_15_Density)  ,TRUE,FALSE))

f3       <- ggplot(filter(GHCN,Umax==TRUE),aes(Pmax, fill=Pmax))+geom_bar()+
            ggtitle("GHCN Stations matching CRN",subtitle = "10 km Population")+
            xlab("CRN-like") + ylab("Station Count")
f3


 
f4       <- ggplot(filter(GHCN,Umax==TRUE & Pmax==TRUE), aes(GPW10km_15_Density))+geom_histogram() +
           ggtitle("Area Population",subtitle = "10 km radius")+
           xlab("Density") + ylab("Station Count")
      
f4
```

The next filter we apply is for Hyde based Population figures
 

```{r, message = FALSE,warning=FALSE}
 Maxhyde <- quantile(CRN$Hyde_density00,probs=seq(0,1,by=.05))[19]
GHCN     <- GHCN %>% mutate(Hmax = ifelse(Hyde_density00 < Maxhyde | is.na(Hyde_density00)  ,TRUE,FALSE))

f4       <- ggplot(filter(GHCN,Umax==TRUE & Pmax==TRUE),aes(Hmax, fill=Hmax))+geom_bar()+
            ggtitle("GHCN Stations matching CRN",subtitle = "Hyde Population")+
            xlab("CRN-like") + ylab("Station Count")
f4


 
f5       <- ggplot(filter(GHCN,Umax==TRUE & Pmax==TRUE & Hmax==TRUE), aes(GPW10km_15_Density))+geom_histogram() +
           ggtitle("Area Population",subtitle = "10 km radius")+
           xlab("Density") + ylab("Station Count")
      
f5

f6 <- ggplot(filter(GHCN,Umax==TRUE & Pmax==TRUE & Hmax==TRUE), aes(100* UrbanArea10K/314))+geom_histogram() +
           ggtitle("Percentage Urban Cover",subtitle = "10 km radius")+
           xlab("Percentage") + ylab("Station Count")
f6
```

```{r, message = FALSE,warning=FALSE}
 Maxgpw <- quantile(CRN$GpwV4_density00,probs=seq(0,1,by=.05))[19]
GHCN     <- GHCN %>% mutate(Gmax = ifelse(GpwV4_density00 < Maxgpw | is.na(GpwV4_density00)  ,TRUE,FALSE))

f7       <- ggplot(filter(GHCN,Umax==TRUE & Pmax==TRUE & Hmax==TRUE),aes(Gmax, fill=Hmax))+geom_bar()+
            ggtitle("GHCN Stations matching CRN",subtitle = "Gpw Population")+
            xlab("CRN-like") + ylab("Station Count")
f7


 
f8       <- ggplot(filter(GHCN,Umax==TRUE & Pmax==TRUE & Hmax==TRUE & Gmax==TRUE), aes(GPW10km_15_Density))+geom_histogram() +
           ggtitle("Area Population",subtitle = "10 km radius")+
           xlab("Density") + ylab("Station Count")
      
f8

f9 <- ggplot(filter(GHCN,Umax==TRUE & Pmax==TRUE & Hmax==TRUE & Gmax==TRUE), aes(100* UrbanArea10K/314))+geom_histogram() +
           ggtitle("Percentage Urban Cover",subtitle = "10 km radius")+
           xlab("Percentage") + ylab("Station Count")
f9
```
## Distance to Airport 
  Some CRN stations are located close to small grass field airports or medium sized single runaway airports.
  One is located roughly 3km from a large airport Ashland Regional Airport. This distance will be applied as a filter to all GHCN  stations and those closer than this distance to any sized airport will be culled.

```{r, message = FALSE,warning=FALSE}
Minap <- min(CRN$Airport_Dist[CRN$Airport_Type=="large_airport"])
GHCN     <- GHCN %>% mutate(Amin = ifelse(Airport_Dist > Minap  ,TRUE,FALSE))

f7       <- ggplot(filter(GHCN,Umax==TRUE & Pmax==TRUE & Hmax==TRUE &Gmax==TRUE),aes(Amin, fill=Amin))+geom_bar()+
            ggtitle("GHCN Stations matching CRN",subtitle = "Distance from Aiport")+
            xlab("CRN-like") + ylab("Station Count")
f7


 
f8       <- ggplot(filter(GHCN,Umax==TRUE & Pmax==TRUE & Hmax==TRUE &  Gmax==TRUE &Amin==TRUE), aes(GPW10km_15_Density))+geom_histogram() +
           ggtitle("Area Population",subtitle = "10 km radius")+
           xlab("Density") + ylab("Station Count")
      
f8

f6 <- ggplot(filter(GHCN,Umax==TRUE & Pmax==TRUE & Hmax==TRUE & Gmax==TRUE & Amin==TRUE), aes(100* UrbanArea10K/314))+geom_histogram() +
           ggtitle("Percentage Urban Cover",subtitle = "10 km radius")+
           xlab("Percentage") + ylab("Station Count")
f6


```

## Landform

In addition to the elevation and the local DEM elevation we also collect the land form of the site which indicates whether the site is on smooth plains, hills, valleys. 

```{r, message = FALSE,warning=FALSE}
f13 <- ggplot(GHCN, aes( EF_LF_Desc,fill=EF_LF_Desc))+geom_bar() +
      theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))+
      ggtitle("Landform")+
      xlab("Landform Type") + ylab("Station Count")

f13

 

```

## Landclass

We use the ESA 300 meter land class  data ( 2008-2012 epoch) to ascertain the land cover for the sites' location. The average user accuracy is ~74%, depending on the land class.

```{r, message = FALSE,warning=FALSE}

S <- GHCN %>% select(Name, LCCOwnLabel) %>%group_by(LCCOwnLabel)%>%
         count()
HighCount <-S$LCCOwnLabel[S$n > median(S$n)]

f14 <- ggplot(filter(GHCN, LCCOwnLabel %in% HighCount), aes( LCCOwnLabel))+geom_bar() +
      coord_flip()+
      ggtitle("Land Cover at the site")+
      xlab("Landform Type") + ylab("Station Count")

f14

f14 <- ggplot(filter(GHCN, !LCCOwnLabel %in% HighCount), aes( LCCOwnLabel))+geom_bar() +
      coord_flip()+
      ggtitle("Land Cover at the site")+
      xlab("Landform Type") + ylab("Station Count")

f14

```

 

## Gridded Population

For every station we extract the population count from GPW  version 4. The gridded data is a 30 arc second dataset or a 1km resolution at the equator. Since grid cell size changes with latitude other equal area representations (10km)are included as well. For GPW4 we inlcude 4 reference dates, 2000, 2005,2010, and 2015.

```{r, message=FALSE,warning=FALSE}
POP <- GHCN %>% select(Name,GPwV4_00,GPwV4_05,GPwV4_10,GPwV4_15) %>%
       gather(Year,Population,starts_with("GP")) %>% mutate(Year=as.factor(Year))  
   
f15<- ggplot(filter(POP, Population < 5000 ),  aes(Population,fill=Year))+geom_histogram()+facet_wrap(~Year,nrow=2,ncol=2) +
      ggtitle("Gridded Population Count", subtitle="Population count Less than 5000")
f15

f16<- ggplot(filter(POP, Population >= 5000),  aes(Population,fill=Year))+geom_histogram()+facet_wrap(~Year,nrow=2,ncol=2) +
      ggtitle("Gridded Population Count", subtitle="Population count greater than 5000")
f16

f17<- ggplot(filter(POP, Population < median(Population,na.rm=TRUE)),  aes(Population,fill=Year))+geom_histogram()+facet_wrap(~Year,nrow=2,ncol=2) +
      ggtitle("Gridded Population Count", subtitle="Population count Less than Median")
f17

f18<- ggplot(filter(GHCN,GPwV4_15_10km < median(GPwV4_15_10km,na.rm=TRUE)),  aes(GPwV4_15_10km))+geom_histogram() +
      ggtitle("Gridded Population Count-less than median",subtitle="10Km radius") 
       
f18

f19<- ggplot(filter(GHCN,GPwV4_15_10km >= median(GPwV4_15_10km,na.rm=TRUE)),  aes(GPwV4_15_10km))+geom_histogram(bins=20) +
      ggtitle("Gridded Population Count- count greater than median",subtitle="10Km radius") 
       
f19

f20<- ggplot(filter(GHCN,GPwV4_15_10km < 100000),  aes(GPwV4_15_10km))+geom_histogram() +
      ggtitle("Gridded Population Count-count less than 100K",subtitle="10Km radius") 
       
f20

f21 <-ggplot(filter(GHCN,GPwV4_15-GPwV4_00 >=0),  aes(GPwV4_15-GPwV4_00 ))+
      geom_histogram(bins=30)+
      ggtitle("Population Growth 2000-2015")+
      xlab("Increase")+ylab("Station Count")
      
f21

f22 <-ggplot(filter(GHCN,GPwV4_15-GPwV4_00 < 0),  aes(GPwV4_15-GPwV4_00 ))+
      geom_histogram(bins=30)+
      ggtitle("Population Decline 2000-2015")+
      xlab("Increase")+ylab("Station Count")
      
f22

```

## Hyde Population

For every station we extract the population count from Hyde  version 3.1. The gridded data is a 5 minute dataset or a 8km resolution at the equator. Since grid cell size changes with latitude other equal area   For Hyde we inlcude 5 reference dates: 1970, 1980,1990, 2000, and 2005.

```{r, message=FALSE,warning=FALSE}
HYDE <- GHCN %>% select(Name,Hyde1970,Hyde1980,Hyde1990,Hyde2000,Hyde2005) %>%
       gather(Year,Population,starts_with("Hy")) %>% mutate(Year=as.factor(Year))
   
f23<- ggplot(filter(HYDE,Population<median(Population,na.rm=TRUE)),  aes(Population,fill=Year))+geom_histogram()+facet_wrap(~Year,nrow=2,ncol=3) +
      theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))+
      ggtitle("Hyde Gridded Population Count-less than median")
f23

f24<- ggplot(filter(HYDE,Population >= median(Population,na.rm=TRUE) & Population < 1000000),  aes(Population,fill=Year))+geom_histogram()+facet_wrap(~Year,nrow=2,ncol=3) +
      theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))+
      ggtitle("Hyde Gridded Population Count-Greater than median")
f24

f25<- ggplot(filter(HYDE,Population > 1000000),  aes(Population,fill=Year))+geom_histogram()+facet_wrap(~Year,nrow=2,ncol=3) +
      theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))+
      ggtitle("Hyde Gridded Population Count-Greater than 1M")
f25
 

f26 <-ggplot(GHCN,  aes(Hyde2005-Hyde1970 ))+
      geom_histogram()+
      ggtitle("Population Growth 1970-2005")+
      xlab("Increase")+ylab("Station Count")
      
f26
```

##  Population Density
 The population density of the site can be calculated by dividing the population count by the area of the cell which is a function of latitute in addition for the 10km population we cans imply divide by the area of a circle with radius 10km. Subtracting these two can give a crude indication of whether a site is more heavily populated at the exact site location or whether the surrounding area has a higher population. 
 
```{r, message=FALSE,warning=FALSE}
 
 f27 <-ggplot(filter(GHCN,GPW10km_15_Density<500),  aes(GPW10km_15_Density ))+
      geom_histogram()+
      ggtitle("Population Density 2015-less 500",subtitle="10km radius")+
      xlab("Population Density per sqkm")+ylab("Station Count")
      
f27

f28 <-ggplot(filter(GHCN,GPW10km_15_Density >=500),  aes(GPW10km_15_Density ))+
      geom_histogram()+
      ggtitle("Population Density 2015-greater than 500",subtitle="10km radius")+
      xlab("Population Density per sqkm")+ylab("Station Count")
      
f28

f29 <-ggplot(GHCN,  aes(GPW10km_15_Density-(GPwV4_15/GPwV4_Area) ))+
      geom_histogram()+
      ggtitle("2015 Population Density Difference",subtitle="10km -Site")+
      xlab("Density Difference")+ylab("Station Count")
      
f29



```

## Isolation from Urban Areas
 In addition the site characteristics and immediate vincity, we also calculate the distance from the site to areas of concentrated population. Several measures are provided. The distance to the closest populated place; the distance to the closest place with a population (circ 2000) greater than 50,000; the distance to the closest place with a population (circ 2000) greater than 500,000;the distance to the closest place with a population (circ 2000) greater than 1 Million;the distance to the closest place with a population (circ 2000) greater than 5M. For reference, "Simulations of the London urban heat island" S. I. Bohnenstengel had results that indicated UHI could advect up to 40km under certain conditions, and Basset ("Observations of urban heat island advection from a high-density monitoring network") indicated advection under certain conditions extending to at least 15km from Birmingham ( Population 1M). Finally, Brandsma ("EMPIRICAL ESTIMATION OF THE EFFECT OF URBAN HEAT ADVECTION ON THE TEMPERATURE SERIES OF DE BILT (THE NETHERLANDS)"), indicates that roughly 10% of the warming seen at De Bilt or .1C out of 1C, is due to advection of UHI from the surrounding cities. Utrecht, the largest city at ~250K, is around 2km  from the site and one smaller town (60K) is ~3km from the site.

```{r, message=FALSE,warning=FALSE}

DIST <- GHCN %>% select(Name,DistanceToPlace50K,DistanceToPlace500K,DistanceToPlace1M,DistanceToPlace5M) %>%
  gather(PlaceSize,Distance, starts_with("DistanceToPlace")) %>% 
  mutate(PlaceSize=str_replace(PlaceSize,"DistanceToPlace","")) %>%
  mutate(PlaceSize=as.factor(PlaceSize)) %>%filter(Distance < 125)

f30 <- ggplot(DIST,  aes(Distance,fill=PlaceSize))+geom_histogram(bins=50)+
       facet_wrap(~PlaceSize,nrow=2,ncol=2,scales="free_y")+
       ylab("Station Count")+
       ggtitle("Isolation From Urban Centers")
f30
```

## Distance to Airports

For every site we calculated the distance to the closests airport. There are three type of airports: small, medium and large. Small and medium size airports are typically single runway installations.

```{r, message= FALSE,warning=FALSE}


 
f31  <-ggplot(GHCN,  aes(Airport_Dist,fill=Airport_Type ))+
  geom_histogram(bins=60,position="stack")+
  ggtitle("Distance to Nearest Airport",subtitle="small medium and large airports")+
  xlab("Distance in km")+ylab("Station Count")

f31

f32  <-ggplot(GHCN,  aes(Airport_Dist2,fill=Airport_Type2 ))+
  geom_histogram(bins=60,position="stack")+
  ggtitle("Distance to Nearest Airport",subtitle="medium and large airports")+
  xlab("Distance in km")+ylab("Station Count")

f32

f33  <-ggplot(filter(GHCN, Airport_Dist < 10),  aes(Airport_Dist,fill=Airport_Type ))+
  geom_histogram(bins=60,position="stack")+
  ggtitle("Distance to Nearest Airport",subtitle="small medium and large airports")+
  xlab("Distance in km")+ylab("Station Count")

f33

f34  <-ggplot(filter(GHCN, Airport_Dist2 < 10),  aes(Airport_Dist2,fill=Airport_Type2 ))+
  geom_histogram(bins=60,position="stack")+
  ggtitle("Distance to Nearest Airport",subtitle="medium and large airports")+
  xlab("Distance in km")+ylab("Station Count")

f34

 
```

## Nightlights

```{r, message = FALSE,warning=FALSE}


f35  <-ggplot(GHCN,  aes(Lights))+
  geom_histogram(bins=60)+
  ggtitle("Radiance Calibrated Nightlights") +
  xlab("Lights")+ylab("Station Count")

f35

f36  <-ggplot(filter(GHCN, Lights <  median(Lights,na.rm=TRUE)),  aes(Lights))+
  geom_histogram(bins=60)+
  ggtitle("Radiance Calibrated Nightlights") +
  xlab("Lights")+ylab("Station Count")

f36

f36 <- ggplot(GHCN, aes(x=GPwV4_10/GPwV4_Area,y=Lights))+geom_point()+
       ggtitle("Lights versus Population Density", subtitle= "circa 2010")+
       xlab("Site Population Density")

f36

```



 